Cell type assignments for SCPCP000015

### 2025-02-04 analyses/cell-type-ewings/exploratory_analysis/08-merged-celltypes.Rmd
incorporate consensus cell types

```{r}
all_info_df <- all_info_df |> 
  dplyr::mutate(
    final_annotation = dplyr::case_when(
      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "ews_low_caf",
      `auc_aynaud-ews-targets` > 0.025 | 
        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") ~ "ews_high",
      .default = consensus_annotation
    )
  )

@@

annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "consensus_lumped", 
  cluster_column = "final_annotation", 
  legend_title = "AUC"
)

# get annotation colors for cell type and cluster 
cell_types <- unique(all_info_df$final_annotation)
num_cell_types <- length(cell_types)
cell_type_colors <- palette.colors(palette = "alphabet") |>
  head(n = num_cell_types) |>
  purrr::set_names(cell_types)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  cell_type = all_info_df$final_annotation,
  col = list(
    cell_type = cell_type_colors
  )
)

# build matrix for heatmap cells x gene set/ expression columns 
heatmap_mtx <- all_info_df |>
  dplyr::select(barcodes, all_of(auc_columns)) |>
  tibble::column_to_rownames("barcodes") |>
  as.matrix() |>
  t()

# remove any extra strings from the expression column names 
rownames(heatmap_mtx) <- stringr::str_remove(rownames(heatmap_mtx), "auc_|_mean")

# plot heatmap of marker genes
plot_gene_heatmap(heatmap_mtx,
                  row_title = "",
                  legend_title = "final_annotation",
                  annotation = annotation,
                  cluster_columns = TRUE
)
```

@@

```{r, fig.height=8}
ggplot(exp_df, aes(x = mean, color = tumor_cell_classification, fill = tumor_cell_classification)) +
  geom_density(alpha = 0.5, bw = 0.5) +
  facet_wrap(vars(geneset)) +
  theme(
    aspect.ratio = 1,
    strip.background = element_rect(fill = "transparent", linewidth = 0.5),
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5)
  )
```

@@

```{r, fig.height=30, warning=FALSE}
mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "consensus_annotation", "mean gene expression")
```

The heatmap below shows the mean expression values for all cells and all gene sets. 
The annotations shown are the `SingleR` cell types and clusters.

```{r}
annotated_exp_heatmap(
  all_info_df, 
  exp_columns = mean_exp_columns, 
  cell_type_column = "singler_lumped", 
  cluster_column = "singler_lumped", 
  legend_title = "Mean\ngene expression"
)
```

## Re-cluster tumor cells {.manual-exploration}

<!-- Use this section to label tumor cells based on the above findings. 
Any cells that are labeled as tumor will then be re-clustered and plots showing only tumor cells
can be created to identify tumor cell states -->

TODO: Functions for re-clustering tumor cells
Show the same plots across the tumor only clusters and assign tumor cell states to each cluster

## Additional exploration {.manual-exploration}

<!--This section is for any additional exploration that may be needed to finalize annotations for this library. 
If not using, please delete this section.
For example, here you may want to dive into the normal cell types and make adjustments as needed
-->

## Validate final tumor and normal annotation {.manual-exploration}

<!-- This section should be used to update the assignments -->

TODO: Insert plots that will be useful for validation (UMAPs, heatmaps, density plots)

## Prepare annotations {.manual-exploration}

<!-- This section should be used to create the final.final table with cell type annotations for export -->

TODO: Code and instructions for exporting such as what columns should be named.

>>>>

```{r}
all_info_df <- all_info_df |> 
  dplyr::mutate(
    final_annotation = dplyr::case_when(
      `auc_wrenn-nt5e-genes` > 0.1 & auc_miyagawa_down > 0.03 ~ "ews_low_caf",
      `auc_aynaud-ews-targets` > 0.025 | 
        consensus_annotation %in% c("Unknown", "muscle cell", "smooth muscle cell") ~ "ews_high",
      .default = consensus_annotation
    ),
    final_lumped = final_annotation |>
          forcats::fct_lump_n(8, other_level = "All remaining cell types", ties.method = "first") |>
          forcats::fct_infreq() |>
          forcats::fct_relevel("All remaining cell types", after = Inf)
  )

@@

annotated_exp_heatmap(
  all_info_df, 
  exp_columns = auc_columns, 
  cell_type_column = "consensus_lumped", 
  cluster_column = "final_lumped", 
  legend_title = "AUC"
)
```

@@

@@

```{r, fig.height=7, warning=FALSE}
mean_exp_columns <- colnames(all_info_df)[which(endsWith(colnames(all_info_df), "_mean"))]
cluster_density_plot(all_info_df, mean_exp_columns, annotation_column = "final_lumped", "mean gene expression")
```

What are the other cell types not present here? 
```{r}
lumped_celltypes <- unique(all_info_df$final_lumped)
all_celltypes <- unique(all_info_df$final_annotation)
missing_celltypes <- setdiff(all_celltypes, lumped_celltypes)

missing_only <- all_info_df |> 
  dplyr::filter(final_annotation %in% missing_celltypes)
table(missing_only$final_annotation)
```

I think we can leave these few cells alone and just lump them as "All remaining cell types" for plots as we have been doing.

Now let's make a dot plot! Or a heatmap only looking at key genes! 
In these plots, lets group all T cells together.

```{r}
markers_df <- readr::read_tsv(file.path(module_base, "references", "combined-markers.tsv"))
genes <- markers_df |>
  dplyr::pull(ensembl_gene_id)

gene_cts <- logcounts(sce[genes, ]) |>
  as.matrix() |>
  t() |> 
  as.data.frame()
gene_cts$barcodes <- rownames(gene_cts)

rowData(sce)[genes, ] |> 
  as.data.frame() |> 
  dplyr::select(ends_with(".mean")) |> 
  tibble::rownames_to_column("ensembl_gene_id") |> 
  tidyr::pivot_longer(!ensembl_gene_id, names_to = "library_id", values_to = "mean")

celltypes_df <- endsWith()celltypes_df <- all_info_df |> 
  dplyr::select(barcodes, final_lumped)

genes_df <- gene_cts |> 
  tidyr::pivot_longer(!barcodes, names_to = "ensembl_gene_id", values_to = "gene_exp") |> 
  dplyr::left_join(celltypes_df, by = c("barcodes")) |> 
  dplyr::mutate(
    percent_expression = 
  )

head(genes_df)
```
